Correlation tests

### 2025-04-04 analysis/manuscript.tex
Rough intro outline

This study aims to demonstrate the suitability of open scholarly data
sources for assessing the impact of transformative agreements on hybrid
open access. To achieve this, a replication study was conducted by
comparing results from hoaddata, an openly available and continuously
updated dataset on hybrid open access uptake based on Crossref,
OpenAlex, and the cOAlition S Journal Checker Tool, with the established
bibliometric databases Web of Science and Scopus.

This study focuses on the coverage of hybrid journal portfolios included
in transformative agreements between 2019 and 2023. Special attention is
given to potential differences in open access uptake by country when
comparing first-author affiliation data to corresponding authorships.
This is crucial because the lack of publicly available invoicing data
corresponding to authorships plays an essential role in determining
whether an open-access article is supported through transformative
agreements. Data on corresponding authorships have been available on the
Web of Science and Scopus for much longer than in open databases such as
OpenAlex, where this information is still being roled out at the time of
writing. Because of this weakness, open approaches such as hoaddata and
related research use first-authorship data instead.

By conducting a large-scale comparative analysis, this study aims to

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Determine the strengths and weaknesses of using open data sources in
  monitoring the impact of transformative agreements on hybrid open
  access publishing.
\item
  Assess the coverage and accuracy of open data sources compared with
  established bibliometric databases.
\item
  Evaluate the reliability of first author affiliation data as a proxy
  for corresponding authorship in the context of open access uptake
  analysis.
\end{enumerate}

\section{Background -- Evidence base to measure the effects of
transformative
agreements}\label{background-evidence-base-to-measure-the-effects-of-transformative-agreements}

\subsection{Anforderungen an das
Monitoring}\label{anforderungen-an-das-monitoring}

\begin{itemize}
\tightlist
\item
  esac guidelines
\item
  gemeinsamkeiten und unterschiede zu apc (listenpreise, tatsächliche
  zahlungen, zentrales invoicing, rabatte, waivers)
\item
  insitutionen covern cas, jedoch kann es zu unterschiedlichen
  verrechnugnsformen führen (antielig mit förderer, splitting
  innerhaklbd er einrichtung)
\end{itemize}

\subsection{Bibliometrische Evidenzen}\label{bibliometrische-evidenzen}

\begin{itemize}
\tightlist
\item
  allgmeeiner uptake
\item
  wachstum apcs
\item
  wachstum verträge (konsortien, forschung)
\item
  konsequenzen
\end{itemize}

@@

This study of over 13,000 hybrid journals included in the Transformative
Agreement Data underlying the cOAlition S Journal Checker Tool shows a
rise in open access due to transformative agreements between 2019 and
2023, although most articles remained paywalled. While transformative
agreements accounted for the majority of open access, many articles
continue to become open through the payment of individual publication
fees. Hybrid open access and transformative agreements remain
concentrated among a small group of large commercial publishers, with
high-income European countries---alongside South Africa---showing high
adoption rates. In contrast, the three most productive countries, China,
the United States, and India, were ranked much behind other countries in
the adoption of transformative agreements. Open questions remain whether
this uneven distribution reflects temporary implementation gaps,
inherent inequities in the transformative agreement model, or countries
voluntarily refraining from such agreements.

These trends were consistent across the investigated open data source
hoaddata, derived from Crossref and OpenAlex, and the established
proprietary bibliometric databases Scopus and Web of Science. Notably,
the results show strong correlations by country affiliation, aligning
with previous studies (Akbaritabar et al., 2024; Alperin et al., 2024),
which supports the use of open research information for large-scale
analyses of hybrid open access. However, the observed differences in
coverage of journals, articles, and metadata availability warrant
discussion, affecting not only open data sources but also proprietary
databases when used in isolation.

To begin with, the coverage analysis reveals that hybrid journals are
well indexed in all three data sources, particularly in terms of article
coverage. Differences can be found for journals targeting practitioners
or local non-English language communities, with many such titles indexed
exclusively in Crossref and Scopus. Using the open index Crossref as a
bibliometric database in hoaddata demonstrated particular strength in
identifying newly established hybrid journals, a notable finding given
that transformative agreements primarily target existing
subscription-based journals.

The situation of hybrid journal publishing is therefore different from
that of fully open access journals. Comparing the coverage of OpenAlex,
Scopus and Web of Science, Simard et al. (2024) indicate that only half
of the fully open access journals listed in the DOAJ are also indexed in
Scopus and Web of Science. Notably, journals that charge no publication
fees (``diamond journals'') are absent from the selective Web of
Science, which reinforces existing disparities in the indexing of
underrepresented research communities and regions in selective
bibliometric databases (ref).

One often reported limitation of studying open access with less
selective databases is the lack of corresponding authorship information
(Fraser et al., 2023; Haucap et al., 2021; Shu \& Larivière, 2023).
However, indicators based on first authors, which have often been used
as a proxy for determining open access funding, and corresponding
authors show a high level of correlation, reflecting disciplinary norms
in scholarly publishing with regard to contributorship and author roles
and positions (Larivière et al., 2016). First authors are often
considered to have carried out the main research underlying a paper,
while the corresponding author is often recognised as having supervised
the research (Fox et al., 2018; Mattsson et al., 2010). Unsurprisingly,
measures based on first or corresponding authorships are strongly
correlated, suggesting that these authors share the same country
affiliation. Moreover, in most cases, the first author is identical the
corresponding author (Chinchilla-Rodríguez et al., 2024). Thus, first
authorship can reliably be used in large-scale studies to assess the
impact of transformative agreements.

Despite this correlation, the recent slight increase in open access
articles by corresponding authors over first authors across Springer
Nature hybrid journals prompts questions about how institutional open
access sponsorship practices influence author roles and assignments
within co-author teams, especially as funding opportunities vary
(Gumpenberger et al., 2018). Previous research has highlighted how
institutionalized bibliometric practices can affect the valuation of
authorship positions (Helgesson, 2020), underscoring the need to be
cautious of monetary authorship influences (Maddi \& Silva, 2024).

Affiliation data is another important data element for estimating open
access enabled by transformative agreements. Although OpenAlex's
affiliation coverage is less comprehensive, which likely reduced the
number of articles confidently attributed to transformative agreements
in this study, it still shows high correlations with Scopus and Web of
Science at the country level. However, OpenAlex's native ROR ID
integration offers a distinct advantage, enabling more reliable
identification of agreement-enabled articles compared to Scopus and Web
of Science, which require reconciliation with proprietary organization
identifiers. Future studies based on the Web of Science will benefit of
the recent integration of ROR-IDs, announced by the end of 2024.

Comparing scholarly databases reveals discrepancies in open access
evidence. Not all publishers deposit Creative Commons (CC) licenses with
Crossref, but comparing it with Unpaywall (integrated into Scopus and
Web of Science) exposes these gaps, as these discovery services also
analyze publishers' websites.

Comparing scholarly databases helps to identify discrepancies in open
access evidence. Crucially, not all publishers deposit CC licences with
Crossref, a limitation that becomes apparent when contrasting Crossref
with Unpaywall's data in Scopus and Web of Science. While Unpaywall can
detect such gaps by parsing publishers' websites for open access
licences, it nevertheless missed a substantial number of CC-licensed
open access articles from Wiley journals indexed in Crossref, likely due
to parsing errors on the publisher's website. This resulted in fewer
open access articles being recorded in Scopus and Web of Science.
Further inconsistencies could be observed between Scopus and Web of
Science, despite both relying on Unpaywall: Web of Science erroneously
labels Elsevier's delayed OA as hybrid, whereas Scopus correctly
categorises it. ONotably, Scopus and Crossref showed greater alignment
than observed in a related comparison between Scopus and OpenAlex
(Alperin et al., 2024). This likely reflects the use of a dedicated list
to identify hybrid open access journals, rather than relying solely on
article-level tags. Given these variations, which may compromise
comparability, research and monitoring exercises should avoid reliance
on a single source. Instead, the selection of open access evidence
should be cross-verified using multiple sources (Huang et al., 2020).
Incorporating expert-curated journal lists can also help prevent
misclassifications based on business models (Visser et al., 2021).

It is important to note that articles from institutions involved in
transformative agreements are estimates due to a lack of access to
invoice data, which is not usually shared between library consortia and
publishers. Furthermore, it is important to acknowledge that the sources
of transformative agreements and the underlying ESAC registry are a
voluntary effort, crowd-sourced from various consortia. However, recent
validation with Dutch Research Information demonstrates the reliability
of such an open approach for the assessment of articles under
transformative agreements. With enhanced and streamlined metadata
compliance, attributable to evolving standards and initiatives to
support negotiations with publishers, particularly through the
communities of the ESAC initiative and Barcelona Declaration supporters,
the situation is likely to improve. In conclusion, this cross-database
analysis has highlighted the potential and challenges of establishing
open scholarly data as a suitable evidence base for analysing the
transition from subscription-based publishing to open access.

@@

>>>>

Transformative agreements have emerged as a much-discussed licensing
model for the transition from subscription-based publishing to full open
access in the last ten years. While these agreements can vary
substantially, increasing coordination exists in standardising the
negotiation, administration, and evaluation of these agreements.
Simultaneously, efforts to improve transparency through open data about
contracts and payments are growing. However, this information remains
only partially available through bibliometric databases, limiting
large-scale studies about the impact of transformative agreements on the
transition of hybrid journals to full open access through increased open
access uptake. This study compares findings from hoaddata, an openly
available, continuously updated dataset on hybrid open access, built on
open data from Crossref, OpenAlex and the cOAlition S Journal Checler
Tool, with the established bibliometric databases Web of Science and
Scopus. The purpose of this comparison is not only to evaluate the
suitability of open metadata for measuring hybrid open access but also
to inform bibliometric research and practice by highlighting the
strengths and weaknesses of each data source.

Since their introduction, transformative agreements have sparked
considerable academic debate. Whilst standardisation activities,
particularly through the OA2020 and ESAC (Efficiency and Standards for
Article Charges) initiatives, have led to improved workflows between
library consortia and publishers, as well as increased open access
publishing, the initial focus on a few large publishers by national
consortia from high-income countries has drawn criticism. Critics have
raised concerns about market concentration, the lack of journal flipping
to full open access models, reduced competition in the publishing
market, and wider gaps between well-resourced and under-resourced
institutions and researchers.

For instance, studying over 700 agreements, Jahn (2025) observed a
strong growth in open access publishing to 15\% in 2022 due to the
varying implementation of transformative agreements across countries,
predominantly driven by three commercial publishers: Elsevier, Springer
Nature, and Wiley. This aligns with the rapid proliferation of
transformative agreements in recent years. Previously, only a minimal
proportion of articles were made open access in hybrid journals
(Piwowar, 2018; Laakso, 2016).

Evaluations conducted by library consortia confirm these trends but draw
different conclusions. For instance, the British Joint Information
Systems Committee's evaluation (Brayman, Devenney et al., 2024) noted
transformative agreements' significant impact on national open access
growth but limited effect on global open access transition, suggesting a
need to reevaluate their strategy. Similar policy reconsiderations
emerged in Norway and Sweden (Holden, Skoie et al., 2023; Widding,
2024). Meanwhile, cOAlition S recommended ending financial support for
these agreements in 2024, but still considers hybrid open access
compliant with funder open access policies. Conversely, Germany's DEAL
consortium extended agreements with major publishers through 2028, while
Colombia established Latin America's first transformative agreements
(Muñoz-Vélez, Pallares et al., 2024). A recent Max Planck Digital
Library study demonstrates how cost savings can be achieved through this
model and how it can be further developed to achieve full open access
while controlling expenditure.

Open databases like OpenAlex, which build upon Crossref, particularly
benefit from these developments. These resources have gained
considerable attention in research evaluation due to their inclusivity,
openness, and comprehensive coverage of publication metadata. Recent
studies {[}REFERENCE NEEDED{]} have demonstrated that open bibliometric
databases increasingly match or exceed the coverage of traditional
commercial databases in many disciplines, whilst providing improved
access to standardised and interlinked open research information.

Against this background, this study evaluates the suitability of open
metadata to measure the impact of transformative agreements on hybrid
open access. Only a few studies have attempted this analysis, primarily
due to data limitations and the need to combine multiple sources.
Transformative agreements present unique measurement challenges because
of their complex workflow requirements, which are not fully reflected in
bibliometric databases. These challenges include varying journal
portfolios across agreements, invoicing procedures specific to
submitting authors, and article caps that limit open access publishing
within agreements.

The present study addresses these gaps by replicating and applying
consistent methodologies across Web of Science, Scopus, and open data
sources. This comparative approach enables the research to overcome the
limitations of any single source whilst systematically assessing the
strengths and weaknesses of different bibliometric data sources for
monitoring transformative agreement outcomes. By triangulating findings
across these platforms, the study aims to provide a more comprehensive
understanding of how transformative agreements influence hybrid open
access publishing patterns within the ongoing scholarly publishing
transition.

@@

This study of over 13,000 hybrid journals shows a rise in open access
due to transformative agreements between 2019 and 2023, although most
articles remained paywalled. While transformative agreements accounted
for the majority of open access, many articles continue to become open
through the payment of individual publication fees. Hybrid open access
and transformative agreements remain concentrated among a small group of
large commercial publishers, with high-income European
countries---alongside South Africa---showing high adoption rates. In
contrast, the three most productive countries, China, the United States,
and India, were substantially ranked lower in transformative agreement
adoption. Open questions remain whether this uneven distribution
reflects temporary implementation gaps, inherent inequities in the
transformative agreement model, or countries deliberately refraining
from such agreements.

The findings were consistent across the investigated open data source
hoaddata, derived from Crossref and OpenAlex, and the established
proprietary bibliometric databases Scopus and Web of Science. Notably,
the results show strong correlations by country affiliation, aligning
with previous studies (Akbaritabar et al., 2024; Alperin et al., 2024)
and rankings (Eck et al., 2024), which supports using open metadata for
large-scale analyses of hybrid open access. However, the observed
differences in journal coverage and metadata availability warrant
discussion, affecting not only open data sources but also proprietary
databases when used in isolation.

The coverage analysis reveals that hybrid journals are well indexed in
all three data sources, particularly in terms of article coverage.
Differences emerge for journals targeting practitioners or local
non-English language communities, with many such titles indexed
exclusively in Crossref and Scopus. Using the open index Crossref as a
bibliometric database in hoaddata demonstrated particular strength in
identifying newly established hybrid journals, a notable finding given
that transformative agreements primarily target existing
subscription-based journals.

The landscape of hybrid journal publishing differs markedly from that of
fully open access journals. Comparing the coverage of OpenAlex, Scopus
and Web of Science, Simard et al. (2024) indicate that only half of the
fully open access journals listed in the DOAJ are also indexed in Scopus
and Web of Science. Notably, journals that charge no publication fees
(``diamond journals'') are absent from the selective Web of Science,
which reinforces existing disparities in the indexing of
underrepresented research communities and regions in selective
bibliometric databases (ref).

A frequently reported limitation of studying open access with less
selective databases is the lack of corresponding authorship information
(Fraser et al., 2023; Haucap et al., 2021; Shu \& Larivière, 2023).
However, this analysis demonstrates that indicators based on first
authors, which have often been used as a proxy for determining open
access funding, and corresponding authors show a high level of
correlation, reflecting disciplinary norms in scholarly publishing with
regard to contributorship and author roles and positions (Larivière et
al., 2016). First authors typically conduct the main research underlying
a paper, while the corresponding author often supervises the research
(Fox et al., 2018; Mattsson et al., 2010). Unsurprisingly, measures
based on first or corresponding authorships are strongly correlated,
suggesting that these authors share the same country affiliation.
Moreover, in most cases, the first author is identical to the
corresponding author (Chinchilla-Rodríguez et al., 2024). Thus, first
authorship can reliably be used in large-scale studies to assess the
impact of transformative agreements.

Despite this correlation, the study observed a recent slight increase in
open access articles by corresponding authors over first authors across
Springer Nature hybrid journals prompts questions about how
institutional open access sponsorship practices influence author roles
and assignments within co-author teams, especially as funding
opportunities vary (Gumpenberger et al., 2018). Previous research has
highlighted how institutionalized bibliometric practices can affect the
valuation of authorship positions (Helgesson, 2020), suggesting the need
to monitor potential influences on authorships (Maddi \& Silva, 2024).

Another critical data element in the study is affiliation data, which is
essential for estimating open access enabled by transformative
agreements. Although OpenAlex's affiliation coverage is less
comprehensive, which likely reduced the number of articles confidently
attributed to transformative agreements in this study, it still shows
high correlations with Scopus and Web of Science at the country level.
However, OpenAlex's native ROR ID integration offers a distinct
advantage, enabling more reliable identification of agreement-enabled
articles compared to Scopus and Web of Science, which require
reconciliation with proprietary organization identifiers. Future studies
based on the Web of Science will benefit from the recent integration of
ROR-IDs, announced by the end of 2024.

The database comparison revealed important discrepancies in open access
evidence. Crucially, not all publishers deposit CC licences with
Crossref, a limitation that becomes apparent when contrasting Crossref
with Unpaywall's data in Scopus and Web of Science. While Unpaywall can
detect such gaps by parsing publishers' websites for open access
licences, it nevertheless missed a substantial number of CC-licensed
open access articles from Wiley journals indexed in Crossref, likely due
to parsing errors on the publisher's website. This resulted in fewer
open access articles being recorded in Scopus and Web of Science.
Further inconsistencies emerged between Scopus and Web of Science,
despite both relying on Unpaywall: Web of Science erroneously labels
Elsevier's delayed open access as hybrid, whereas Scopus correctly
categorises it. Notably, Scopus and Crossref showed greater alignment
than observed in a related comparison between Scopus and OpenAlex
(Alperin et al., 2024). This likely reflects the use of a curated list
to identify hybrid open access journals, rather than relying solely on
article-level tags. Given these variations, which may compromise
comparability, research and monitoring exercises should avoid reliance
on a single source. Instead, the selection of open access evidence
should be cross-verified using multiple sources (Huang et al., 2020).
Incorporating expert-curated journal lists can also help prevent
misclassifications based on business models (Visser et al., 2021).

It is important to note that the study's estimates of articles from
institutions involved in transformative agreements are approximations
due to a lack of access to invoice data, which is not usually shared by
library consortia and publishers. Furthermore, the data sources about
transformative agreements including the cOAlition S Journal Checker Tool
and the ESAC registry are a voluntary effort, crowd-sourced from various
consortia. However, recent validation with Dutch Research Information
demonstrates the reliability of such an open approach for the assessment
of articles under transformative agreements (Jonge et al., 2025). With
enhanced and streamlined metadata compliance, attributable to evolving
standards and initiatives to support negotiations with publishers,
particularly through the communities of the ESAC initiative and
Barcelona Declaration supporters, the situation is likely to improve. In
conclusion, this study has shown that open scholarly data are suitable
for analysing the transition from subscription-based publishing to full
open access.

@@

\bibitem[\citeproctext]{ref-eck_2024_6bx73}
Eck, N. J. van, Waltman, L., \& Neijssel, M. (2024). \emph{{Launch of
the CWTS Leiden Ranking Open Edition 2024}}. Leiden Madtrics.
\url{https://doi.org/10.59350/r512t-r8h93}
